### 2023-03-02 chapters/background.tex
shortening

\chapter{Preliminaries}
\label{chapter:preliminaries}
We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and the valuation or an assignment of a random variable in lowercase.

@@

\subsection{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
%Let $\boolset  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
%A propositional formula $\propformula$ defined over $\boolset$ is in \textit{Conjunctive Normal Form} (CNF) if $\propformula$   is  a conjunction of clauses and each clause is a disjunction of literals. 
%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
%Let $ \sigma $ be a assignment to the  variables $ \bool_i \in \boolset $  such that $ \sigma(\bool_i) \in \{1, 0\} $. The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \boolset $ such that the formula $ \propformula $ is evaluated to be $1$ (equivalently, true). 
%In contrast to the SAT problem,

The stochastic Boolean satisfiability (SSAT) problem~\cite{littman2001stochastic} is a counting analog of the SAT problem concerning with the probability of satisfaction of the formula. The SSAT problem computes the  probability of satisfaction of a CNF formula $\propformula$ defined on the ordered set of \textit{quantified} Boolean variables $ \boolset $. 
An SSAT formula is of the form
\begin{equation}\label{fairness_justicia_eq:ssat}
	\ssatformula = \varquantifier_1\bool_1, \dots, \varquantifier_m \bool_m,\; \propformula, 
\end{equation}
where $ \varquantifier_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\propformula$ is a quantifier-free CNF formula. In case of a randomized quantifier $ \R^{p_i} $, $ p_i \triangleq \Pr[\bool_i = 1] \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $. In the SSAT formula $ \ssatformula $, the quantifier part $ \varquantifier_1\bool_1, \dots, \varquantifier_m \bool_m $ is known as the \textit{prefix} of the formula $ \propformula $.  Let $ \bool $ be the outermost variable in the prefix. The semantics of SSAT formulas are defined recursively in the following.

@@

\subsection{Stochastic Subset Sum Problem} 
Let $ \boolset \triangleq \{\bool_i\}_{i=1}^{|\boolset|}$ be a set of Boolean variables and $ \varweight_i \in \mathbb{Z} $ be the weight of $ \bool_i $. Given a constraint of the form  $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\boolset \in \{0,1\}^{|\boolset|}$ such that the constraint evaluates to true when $\boolset$ is substituted with $ \mathbf{b} $. Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}. The \textit{counting} version of the subset-sum problem counts all $  \mathbf{b}  $'s for which the above constraint holds. In this thesis, we consider the constraint $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge \tau $ where variables $ \bool_i $'s are stochastic. While the counting problem of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i =  \tau $ is NP-hard to approximate, there is a FPRAS solution for approximately counting the solutions of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge  \tau $~\cite{dyer2003approximate}. In this thesis, we are interested in computing the probability  $ \Pr[\sum_{i = 1}^ {|\boolset|} \varweight_i\bool_i \ge \tau] $, referred to as \textit{stochastic subset-sum}  problem ({\stochastic}).    Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.

>>>>

\chapter{Preliminaries}
\label{chapter:preliminaries}
We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and the valuation or an assignment of a random variable in lowercase. For example, $ \mathbf{X} = \{X_i\} $ is a set of random variables. The distribution of a random variable is $ X \in \mathcal{X} $ and its valuation  is $ X = x $.

@@

\subsection{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
%Let $\boolset  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
%A propositional formula $\propformula$ defined over $\boolset$ is in \textit{Conjunctive Normal Form} (CNF) if $\propformula$   is  a conjunction of clauses and each clause is a disjunction of literals. 
%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
%Let $ \sigma $ be a assignment to the  variables $ \bool_i \in \boolset $  such that $ \sigma(\bool_i) \in \{1, 0\} $. The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \boolset $ such that the formula $ \propformula $ is evaluated to be $1$ (equivalently, true). 
%In contrast to the SAT problem, 
Stochastic Boolean satisfiability (SSAT) was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}.  SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.

SSAT problem~\cite{littman2001stochastic} is a counting analog of the SAT problem concerning with the probability of satisfaction of the formula. SSAT problem computes the  probability of satisfaction of a CNF formula $\propformula$ defined on the ordered set of \textit{quantified} Boolean variables $ \boolset $. 
An SSAT formula is of the form
\begin{equation}\label{fairness_justicia_eq:ssat}
	\ssatformula = \varquantifier_1\bool_1, \dots, \varquantifier_m \bool_m,\; \propformula, 
\end{equation}
where $ \varquantifier_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\propformula$ is a quantifier-free CNF formula. In case of a randomized quantifier $ \R^{p_i} $, $ p_i \triangleq \Pr[\bool_i = 1] \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $. In the SSAT formula $ \ssatformula $, the quantifier part $ \varquantifier_1\bool_1, \dots, \varquantifier_m \bool_m $ is known as the \textit{prefix} of the formula $ \propformula $.  Let $ \bool $ be the outermost variable in the prefix. The semantics of SSAT formulas are defined recursively in the following.

@@

\subsection{Stochastic Subset Sum Problem} 
Let $ \boolset \triangleq \{\bool_i\}_{i=1}^{|\boolset|}$ be a set of Boolean variables and $ \varweight_i \in \mathbb{Z} $ be the weight of $ \bool_i $. Given a constraint of the form  $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\boolset \in \{0,1\}^{|\boolset|}$ such that the constraint evaluates to true when $\boolset$ is substituted with $ \mathbf{b} $. Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}. The \textit{counting} version of the subset-sum problem counts all $  \mathbf{b}  $'s for which the above constraint holds. In this thesis, we consider the constraint $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge \tau $ where variables $ \bool_i $'s are stochastic. While the counting problem of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i =  \tau $ is NP-hard to approximate, there is a FPRAS solution for approximately counting the solutions of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge  \tau $~\cite{dyer2003approximate}. We, in particular, are interested in computing the probability  $ \Pr[\sum_{i = 1}^ {|\boolset|} \varweight_i\bool_i \ge \tau] $, referred to as \textit{stochastic subset-sum}  problem ({\stochastic}).    Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.

### 2023-03-02 chapters/fairness/fvgm/introduction.tex
shortening

	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} as our first step, because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. In this context, existing approaches suffer from limited scalability while verifying linear classifiers. This is due to the encoding of linear classifiers into SSAT (Chapter~\ref{chapter:justicia}) or SMT formulas \cite{albarghouthi2017fairsquare}. For example, {\justicia} applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.

>>>>

	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. In this context, existing approaches suffer from limited scalability while verifying linear classifiers. This is due to the encoding of linear classifiers into SSAT (Chapter~\ref{chapter:justicia}) or SMT formulas \cite{albarghouthi2017fairsquare}. For example, {\justicia} applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.

### 2023-03-02 chapters/fairness/justicia/introduction.tex
shortening

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
\label{chapter:justicia}
Machine learning is becoming the omnipresent technology of our time. Machine learning classifiers are being used for high-stake decisions such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. Thus, human lives are now pervasively influenced by data, classifiers, and their inherent bias.

\begin{example}\label{fairness_justicia_example:intro}
\normalfont
Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60). Typically, incomes of individuals increase as their ages increase while their fitness deteriorates (Figure~\ref{fairness_justicia_fig:fair_example_dependency}). We assume that the relation of income and fitness depends on ages as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example_distribution}. Now, if we train a decision tree~\cite{narodytska2018learning} to decide the eligibility of an individual to get a health insurance given three features: fitness, income and age, we observe that the `optimal' decision tree (ref. Figure~\ref{fairness_justicia_fig:fair_example_dt}) does not predict based on the sensitive feature age. However, a fairness verifier would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively (Figure~\ref{fairness_justicia_fig:fair_example_probability}). This simple example demonstrates that even if a classifier does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
\end{example}

@@

\paragraph{Fairness in Machine Learning.} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias. In this chapter, we focus on a popular concept of fairness, known as group fairness. Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}. Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}. Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth class label~\cite{hardt2016equality}. Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.

In Figure~\ref{fairness_justicia_fig:fair_example}, independence is satisfied if the probability of getting insurance is same for both the age groups. Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same. Sufficiency is satisfied if the eligibility is independent of their age given their features are the same. Thus, we see that the metrics of fairness can be contradictory and complementary depending on the application and the data~\cite{corbett2018measure}. To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions. These algorithms try to rectify and mitigate the bias in the data and thus in the classifier in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.

\paragraph{Fairness Verifiers.} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.

In order to verify fairness as a model property on a dataset, verifiers like \textit{FairSquare}~\cite{albarghouthi2017fairsquare} and \textit{VeriFair}~\cite{bastani2019probabilistic} have been proposed. 
%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
%These papers operate for a specific Boolean sensitive feature.
These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}. 
Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.

%\blue{Other papers: Probabilistic Verification of Fairness Properties via	Concentration, Verifying Individual Fairness in Machine Learning Models~\cite{john2020verifying}}
Thus, a \textit{unified formal framework} to verify \textit{different fairness metrics} of a classifer, which is \textit{scalable}, capable of \textit{handling compound protected groups}, \textit{robust} with respect to the test data, and \textit{operational on real-life} datasets and fairness-enhancing algorithms, is missing in the literature.

\paragraph{Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}. SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}. In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}. Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

Our contributions are summarised below:
\begin{itemize}
	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group and causal fairness metrics for different datasets and classifiers.
	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
	\item Unlike previously proposed formal probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
%	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
\end{itemize}

It is worth remarking that significant advances in artificial intelligence bear testimony to the right choice of formulation, for example, formulation of planning as satisfiability (SAT)~\cite{kautz1992planning}. In this context, we view that formulation of fairness as SSAT has potential to spur future work from both the modeling and encoding perspective as well as core algorithmic improvements in the underlying SSAT solvers.

>>>>

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
\label{chapter:justicia}
In this chapter, we study formal verification of fairness in machine learning, where we verify the bias of a classifier given the probability distribution of features. To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed. 
%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
%These papers operate for a specific Boolean sensitive feature.
%These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}. 
Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.

%\blue{Other papers: Probabilistic Verification of Fairness Properties via	Concentration, Verifying Individual Fairness in Machine Learning Models~\cite{john2020verifying}}
Thus, a \textit{unified formal framework} to verify \textit{different fairness metrics} of a classifer, which is \textit{scalable}, capable of \textit{handling compound sensitive groups}, \textit{robust} with respect to the test data, and \textit{operational on real-life} datasets and fairness-enhancing algorithms, is missing in the literature.

\paragraph{Contribution.} We propose to model verifying different fairness metrics as a SSAT problem.  We primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.   Our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

Our contributions are summarised below:
\begin{itemize}
	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group  fairness metrics for different datasets and classifiers.
	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
	\item Unlike previously proposed formal probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
%	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
\end{itemize}

%It is worth remarking that significant advances in artificial intelligence bear testimony to the right choice of formulation, for example, formulation of planning as satisfiability (SAT)~\cite{kautz1992planning}. In this context, we view that formulation of fairness as SSAT has potential to spur future work from both the modeling and encoding perspective as well as core algorithmic improvements in the underlying SSAT solvers.

We illustrate the contribution of this chapter using an example scenario.

\begin{example}\label{fairness_justicia_example:intro}
	\normalfont
	Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60). Typically, incomes of individuals increase as their ages increase while their fitness deteriorates (Figure~\ref{fairness_justicia_fig:fair_example_dependency}). We assume that the relation of income and fitness depends on ages as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example_distribution}. Now, if we train a decision tree~\cite{narodytska2018learning} to decide the eligibility of an individual to get a health insurance given three features: fitness, income and age, we observe that the `optimal' decision tree (ref. Figure~\ref{fairness_justicia_fig:fair_example_dt}) does not predict based on the sensitive feature age. However, a fairness verifier, such as {\justicia}, would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively (Figure~\ref{fairness_justicia_fig:fair_example_probability}). This simple example demonstrates that even if a classifier does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
\end{example}

@@

### 2023-03-02 chapters/interpretability/imli/introduction.tex
minor edit
shortening

\label{chapter:imli}
%Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}. In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}. 
In this chapter, we propose an efficient learning framework for interpretable rule-based classifiers. In rule-based classifiers, rules governing the prediction are explicit (by design) in contrast to black-box classifiers~\cite{rudin2019stop}. 
%In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}. Such models are used to either learn an interpretable model from the start or as proxies to provide post-hoc explanations of pre-trained black-box models~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}.  In our context of rule-based classifiers, we use the sparsity of rules as a measure of interpretability, which is adopted in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.  
In particular, we study classifiers expressed as Boolean formulas, wherein we define interpretability in terms of the number of Boolean literals that the formula contains. Here, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.

@@

The problem of learning rule-based classifiers is known to be computationally intractable. The earliest tractable approaches for classifiers such as decision trees and decision lists relied on heuristically chosen objective functions and greedy algorithmic techniques~\cite{ClarkN1989,CohenS1999,quinlan2014}. In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc. In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}. Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, and Bayesian optimizations~\cite{letham2015interpretable} are then deployed to find the best classifier with maximum accuracy and minimum rule-size. In our study, we study an alternate optimization approach that fits particularly well to rule-learning problems. Particularly,  we study and improve a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers. Built on MaxSAT, ~\cite{malioutov2018mlic} considers a learning framework to jointly optimize the accuracy and the size of classification rules. The said approach constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples. Naturally, the na\"ive formulation cannot scale to large values of $ n $ and $ k $. To scale learning to large datasets, we propose an incremental learning technique along with the MaxSAT formulation.

>>>>

\label{chapter:imli}
%Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}. In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}. 
In this chapter, we propose an efficient learning framework for interpretable rule-based classifiers with a particular emphasis on the scalability in learning. In rule-based classifiers, rules governing the prediction are explicit (by design) in contrast to black-box classifiers~\cite{rudin2019stop}. 
%In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}. Such models are used to either learn an interpretable model from the start or as proxies to provide post-hoc explanations of pre-trained black-box models~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}.  In our context of rule-based classifiers, we use the sparsity of rules as a measure of interpretability, which is adopted in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.  
In particular, we study classifiers expressed as Boolean formulas, wherein we define interpretability in terms of the number of Boolean literals that the formula contains. Here, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.

@@

The problem of learning rule-based classifiers is known to be computationally intractable. The earliest tractable approaches for classifiers such as decision trees and decision lists relied on heuristically chosen objective functions and greedy algorithmic techniques~\cite{ClarkN1989,CohenS1999,quinlan2014}. In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc. In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}. Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, and Bayesian optimizations~\cite{letham2015interpretable} are then deployed to find the best classifier with maximum accuracy and minimum rule-size. In our study, we study an alternate optimization approach that fits particularly well to rule-learning problems. Particularly,  we study and improve a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers. Building on MaxSAT, ~\cite{malioutov2018mlic} considers a learning framework to jointly optimize the accuracy and the size of classification rules. The said approach constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples. Naturally, the na\"ive formulation cannot scale to large values of $ n $ and $ k $. To scale learning to large datasets, we propose an incremental learning technique along with the MaxSAT formulation.

### 2023-03-02 chapters/interpretability/relaxed-cnf/experiment.tex
shortening

\section{Experiments}\label{interpretability_crr_sec:experiment}
We implement a prototype  of  $ \crr $ based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances. The objective of our experimental evaluation is to answer the following questions:

@@

%%\vspace{-3pt}
\subsection{Experiment Methodology}
We perform experiments on a high-performance computer cluster, where each node consists of E$ 5$-$2690\text{ v}3 $ CPU with $ 24 $ cores, $ 96 $ GB of RAM. Each experiment is run on four cores of a node with $ 16 $ GB memory.   We compare the performance of  $ \crr $   with  state-of-the-art classifiers, e.g. $ \IMLI  $ (Chapter~\ref{chapter:imli}), RIPPER~\cite{cohen1995fast}, BRS~\cite{wang2017bayesian},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR). Among them, {\IMLI}, BRS, and RIPPER are rule-based classifiers. In particular,
{\IMLI} generates classification rules in CNF using a MaxSAT-based formulation and we use Open-WBO \cite{martins2014open} as the MaxSAT solver for {\IMLI}. We compare with propositional rule learning algorithm RIPPER, which is implemented in WEKA~\cite{HFHPRW2009} and generates classification rules in the form of decision lists. BRS is a Bayesian framework for generating rule sets expressed as DNF. For  other classifiers, we use the Scikit-learn module of Python~\cite{scikit-learn}. For all  classifiers, we set the training cut-off time to $ 1500 $  {seconds}.

@@

%%\vspace{-9pt}
\subsection{Results}
In the following, we first discuss empirical results of rule-based classifiers, then extend analysis to non-rule-based classifiers, and finally discuss the effect of different choices of hyper-parameters. 
%%\vspace{-3pt}
\subsubsection*{Performance Evaluation of {\crr} with Rule-based Classifiers:}
We conduct an assessment of performance  using five-fold nested cross-validation as in~\cite{DGW2018}  and report the median of test accuracy, rule-size and training time of all rule-based classifiers in Table~\ref{interpretability_crr_tab:rule_based_classifiers}. 
Specifically, we show the dataset,  the number of samples and the number of discretized features in the first three columns in Table~\ref{interpretability_crr_tab:rule_based_classifiers}. Inside each cell of column   four to $    11 $,  we present the   test accuracy (top value), rule-size (middle value) and training time (bottom value) of each classifier for each dataset.

@@

\subsubsection*{Performance Evaluation of {\crr} with Non-rule-based Classifiers:}

@@

	\subsubsection*{Varying Model Parameters: }
	\label{interpretability_crr_sec:model_parameters}
	In Figure~\ref{interpretability_crr_fig:result_lambda},~\ref{interpretability_crr_fig:result_clause}, ,~\ref{interpretability_crr_fig:result_minibatch}, and~\ref{interpretability_crr_fig:result_iteration}, we demonstrate the effect of varying the hyper-parameters of {\crr}. To understand the effect of a single hyper-parameter, we fix the values of other hyper-parameters to a default choice where the default choice results in the most accurate rule.

>>>>

\section{Empirical Performance Analysis}\label{interpretability_crr_sec:experiment}
We implement a prototype  of  $ \crr $ based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances. The objective of our experimental evaluation is to answer the following questions:

@@

%%\vspace{-3pt}
\subsection{Experimental Setup}
We perform experiments on a high-performance computer cluster, where each node consists of E$ 5$-$2690\text{ v}3 $ CPU with $ 24 $ cores, $ 96 $ GB of RAM. Each experiment is run on four cores of a node with $ 16 $ GB memory.   We compare the performance of  $ \crr $   with  state-of-the-art classifiers, e.g. $ \IMLI  $ (Chapter~\ref{chapter:imli}), RIPPER~\cite{cohen1995fast}, BRS~\cite{wang2017bayesian},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR). Among them, {\IMLI}, BRS, and RIPPER are rule-based classifiers. In particular,
{\IMLI} generates classification rules in CNF using a MaxSAT-based formulation and we use Open-WBO \cite{martins2014open} as the MaxSAT solver for {\IMLI}. We compare with propositional rule learning algorithm RIPPER, which is implemented in WEKA~\cite{HFHPRW2009} and generates classification rules in the form of decision lists. BRS is a Bayesian framework for generating rule sets expressed as DNF. For  other classifiers, we use the Scikit-learn module of Python~\cite{scikit-learn}. For all  classifiers, we set the training cut-off time to $ 1500 $  {seconds}.

@@

%%\vspace{-9pt}
\subsection{Experimental Results}
In the following, we first discuss empirical results of rule-based classifiers, then extend analysis to non-rule-based classifiers, and finally discuss the effect of different choices of hyper-parameters. 
%%\vspace{-3pt}
\subsubsection{Performance Evaluation of {\crr} with Rule-based Classifiers}
We conduct an assessment of performance  using five-fold nested cross-validation as in~\cite{DGW2018}  and report the median of test accuracy, rule-size and training time of all rule-based classifiers in Table~\ref{interpretability_crr_tab:rule_based_classifiers}. 
Specifically, we show the dataset,  the number of samples and the number of discretized features in the first three columns in Table~\ref{interpretability_crr_tab:rule_based_classifiers}. Inside each cell of column   four to $    11 $,  we present the   test accuracy (top value), rule-size (middle value) and training time (bottom value) of each classifier for each dataset.

@@

\subsubsection{Performance Evaluation of {\crr} with Non-rule-based Classifiers}

@@

	\subsubsection{Varying Model Parameters}
	\label{interpretability_crr_sec:model_parameters}
	In Figure~\ref{interpretability_crr_fig:result_lambda},~\ref{interpretability_crr_fig:result_clause}, ,~\ref{interpretability_crr_fig:result_minibatch}, and~\ref{interpretability_crr_fig:result_iteration}, we demonstrate the effect of varying the hyper-parameters of {\crr}. To understand the effect of a single hyper-parameter, we fix the values of other hyper-parameters to a default choice where the default choice results in the most accurate rule.
